---
title: "appendix"
author: "Taymaz Ghaneh, Maryam Loni, Xixi Zheng"
date: "2025-03-30"
output: pdf_document
---

```{r}
library(dagitty)
library(ggplot2)
library(GGally)
library(gridExtra)
library(cobalt)
library(survey)
library(mice)
library(survey)
library(dagitty)
library(dplyr)
library(devtools)
library(mice)
library(reshape2)
library(BaylorEdPsych)
library(VIM)
library(naniar)
library(stdReg)
```

```{r}
setwd('C:/Users/zxx/OneDrive/Desktop/yr1sem2/causal/group assignment')
data = data.frame(read.csv('2025_framingham_assignment.csv'))
data$HYPERTEN = 0
data[data$SYSBP >= 140 | data$DIABP>= 90 , 'HYPERTEN'] = 1
```

# DAGs

1. Build the first DAG - "g.postulated", according to expertise knowledge.

```{r}
g.postulated <- dagitty('dag {
  SEX [pos="-1, 2"]
  AGE [pos="-3.7, 0.5"]
  CURSMOKE [pos="-0.5, 0.4"]
  DIABETES [pos="-3, 1.9"]
  BPMEDS [pos="-2.5, 0"]
  BMI [pos="-4, 1.3"]
  HEARTRTE [pos="0, 1"]
  HYPERTEN [pos="-2, 1.2"]
  EDUC [pos="-0.3,1.48"]
  
  SEX -> { CURSMOKE BMI HEARTRTE DIABETES HYPERTEN }
  AGE  -> { CURSMOKE DIABETES BMI HYPERTEN BPMEDS EDUC }
  CURSMOKE  -> { BMI HEARTRTE HYPERTEN BPMEDS }
  DIABETES -> { BMI HEARTRTE HYPERTEN BPMEDS }
  BPMEDS -> HYPERTEN
  BMI -> { BPMEDS HYPERTEN }
  HEARTRTE -> { BPMEDS HYPERTEN }
  EDUC -> { BMI CURSMOKE }
  }
')
 
plot(g.postulated, layout = "circle") 
```

2. Find out missing edges in the first DAG and add it -> "g.amended".

```{r}
d = na.omit(data)
r = localTests(g.postulated, d)
r = r[p.adjust(r$p.value) < 0.05, ]
r = r[order(r$p.value), ]
par(mar = c(4,6,4,0))
#plotLocalTestResults(r, bty = "n", xlim = c(-0.4, 0.4), ylim = c(0.1, 3.1),
#                    axis.pars = list(las = 1, lty = 0, cex.axis = 0.6))
```

```{r}
g.amended <- dagitty('dag {
  SEX [pos="-1, 2"]
  AGE [pos="-3.7, 0.5"]
  CURSMOKE [pos="-0.5, 0.4"]
  DIABETES [pos="-3, 1.9"]
  BPMEDS [pos="-2.5, 0"]
  BMI [pos="-4, 1"]
  HEARTRTE [pos="0, 1"]
  HYPERTEN [pos="-2, 1.2"]
  EDUC [pos="-0.3,1.48"]
  
  SEX -> { CURSMOKE DIABETES BMI HEARTRTE HYPERTEN BPMEDS }
  AGE  -> { CURSMOKE DIABETES HEARTRTE HYPERTEN BPMEDS EDUC BMI }
  CURSMOKE  -> { BMI HEARTRTE HYPERTEN BPMEDS }
  DIABETES -> { BMI HEARTRTE HYPERTEN BPMEDS }
  BPMEDS -> HYPERTEN
  BMI -> { BPMEDS HYPERTEN HEARTRTE }
  HEARTRTE -> { BPMEDS HYPERTEN }
  EDUC -> { CURSMOKE BMI HEARTRTE HYPERTEN }
  }
')

plot(g.amended, layout = "circle") 
r1 = localTests(g.amended, d) # no significant relationship
r1 = r1[p.adjust(r1$p.value) < 0.05, ]
r1 = r1[order(r1$p.value), ]
```

3. Find the confounders

```{r}
adjustmentSets(g.amended, exposure = "CURSMOKE", outcome = "HYPERTEN",type = 'all')
```

# Missing data

1. Generate a matrix of missing data patterns and visualize missing patterns.

```{r}
#install_github("alagerberg/BaylorEdPsych")
md.pattern(data)
md_data <- md.pattern(data, plot = FALSE) 
md_data <- md_data[, -ncol(md_data)]
md_df <- as.data.frame(md_data)  

# Convert row names into a column
md_df$Pattern <- rownames(md_df)
md_long <- melt(md_df, id.vars = "Pattern")

ggplot(md_long, aes(x = variable, y = Pattern, fill = value)) +
  geom_tile(color = "black") +
  scale_fill_gradient(low = "white", high = "blue") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +  # Rotate labels
  labs(title = "Missing Data Pattern", x = "Variables", y = "Patterns")
```

2. LittleMCAR is used to test whether the missing pattern of data is MCAR

not MCAR -> use likelihood based method

```{r}
LittleMCAR(data)
```

# description analysis and testing of hypothesis

1. First we take a look at the proportion cross-table between the exposure and the income

```{r CrossTable}
tab = table(data$CURSMOKE, data$HYPERTEN)
round(prop.table((tab), 1),4)
```
2. Check positivity

```{r Pos,fig.width=3, fig.height=3, out.width="50%", out.height="50%"}
par(mfrow=c(2,2))
library(ggplot2)
library(gridExtra)

a = 6 ;  b = 8 ;  c = 8

density1 = ggplot(data, aes(x=AGE, fill=factor(CURSMOKE))) + 
  geom_density(alpha=0.5) +
  labs(title="AGE by CURSMOKE")+
  theme( legend.position = "none",
    plot.title = element_text(size = a),  # Title font size
    axis.title.x = element_text(size = b),  # X-axis label font size
    axis.title.y = element_text(size = c) )  # Y-axis label font size (if needed)

grid.arrange(density1, ncol = 1, nrow=1)
```

3. check exchangeability

```{r Exc, fig.width=8, fig.height=2, out.width="80%", out.height="30%"}
# Proportion of Smokers by Age Group
bar1 = ggplot(data, aes(x = cut(AGE, 
                         breaks = c(30, 40, 50, 60, 70), 
                         labels = c("30-40", "40-50", "50-60", "60-70"), 
                         include.lowest = TRUE)
                 , fill = factor(CURSMOKE))) +
  geom_bar(position = "fill") +  
  labs(title = "", 
       x = "Age Group", 
       y = "Smoker(blue)/non-Smoker(pink) \n Percentage") +
  scale_fill_manual(values = c("pink","steelblue"), 
                    name = "Smoking Status",
                    labels = c("Non-Smoker", "Smoker")) +
  scale_y_continuous(labels = scales::percent)  +
  theme( legend.position = "none",
         plot.title = element_text(size = a),  # Title font size
         axis.title.x = element_text(size = b),  # X-axis label font size
         axis.title.y = element_text(size = c) ,
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(),
         axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))


# Proportion of Smokers by Diabetes
bar2 = ggplot(data, aes(x = factor(DIABETES), fill = factor(CURSMOKE))) +
  geom_bar(position = "fill") +  
  labs(title = "", 
       x = "Diabetes", 
       y = "") +
  scale_fill_manual(values = c("pink","steelblue"), 
                    name = "Smoking Status",
                    labels = c("Non-Smoker", "Smoker")) +
  scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +  # Custom labels
  scale_y_continuous(labels = scales::percent) +
  theme( legend.position = "none",
         plot.title = element_text(size = a),  # Title font size
         axis.title.x = element_text(size = b),  # X-axis label font size
         axis.title.y = element_text(size = c) ,
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(),
         axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

# Proportion of Smokers by Education
bar3 = ggplot(data, aes(x = factor(EDUC), fill = factor(CURSMOKE))) +
  geom_bar(position = "fill") +  
  labs(title = "", 
       x = "Education", 
       y = "") +
  scale_fill_manual(values = c("pink","steelblue"), 
                    name = "Smoking Status",
                    labels = c("Non-Smoker", "Smoker")) +
  scale_x_discrete(labels = c("0" = "No", "1" = "Yes")) +  # Custom labels
  scale_y_continuous(labels = scales::percent) +
  theme( legend.position = "none",
         plot.title = element_text(size = a),  # Title font size
         axis.title.x = element_text(size = b),  # X-axis label font size
         axis.title.y = element_text(size = c) ,
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank(),
         axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))


# Proportion of Smokers by Sex
bar4 = ggplot(data, aes(x = factor(SEX), fill = factor(CURSMOKE))) +
  geom_bar(position = "fill") +  
  labs(title = "", 
       x = "Sex", 
       y = "") +
  scale_fill_manual(values = c("pink","steelblue"), 
                    name = "Smoking Status",
                    labels = c("Non-Smoker", "Smoker")) +
  scale_x_discrete(labels = c("1" = "Male", "2" = "Female")) +  # Custom labels
  scale_y_continuous(labels = scales::percent)  +
  theme( legend.position = "none",
         plot.title = element_text(size = a),  # Title font size
         axis.title.x = element_text(size = b),  # X-axis label font size
         axis.title.y = element_text(size = c) ,
         panel.grid.major = element_blank(),
         panel.grid.minor = element_blank() ,
         axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1) )

grid.arrange(bar1, bar2, bar3, bar4,  ncol = 4, nrow=1)
```

4. Check the balance (exchangeability) after IPW

```{r LOVE, message = FALSE, warning = FALSE, fig.width=5, fig.height=2, out.width="80%", out.height="30%"}
# Drawing LOVE Plot:
library(cobalt)
library(survey)

# Define confounder set
confounders <- c("EDUC", "AGE", "SEX", "DIABETES")

# Function to estimate ACE using IPW
estimate_ACE <- function(confounders, plot_title) {
  
  # Step 1: Remove rows with missing values in CURSMOKE or any of the confounders
  data_clean <- na.omit(data[, c("CURSMOKE", confounders, "HYPERTEN")])  # Assuming HYPERTEN is the outcome variable
  
  # Step 2: Estimate propensity scores using glm
  ps_model <- glm(CURSMOKE ~ ., data = data_clean[, c("CURSMOKE", confounders)], family = binomial())
  data_clean$propensity_score <- fitted.values(ps_model)
  
  # Step 3: Calculate inverse probability weights (IPW)
  data_clean$weights <- ifelse(data_clean$CURSMOKE == 1, 1 / data_clean$propensity_score, 1 / (1 - data_clean$propensity_score))
  
  # Step 4: Estimate ATE using weighted logistic regression (glm)
  ace_model <- glm(HYPERTEN ~ CURSMOKE, data = data_clean, weights = weights, family = binomial())
  
  # Step 5: Extract ATE (log odds difference)
  ATE <- coef(ace_model)["CURSMOKE"]
  
  # Convert ATE from log odds to risk difference (optional)
  ATE_rd <- plogis(coef(ace_model)["CURSMOKE"]) - plogis(0)
  
  # Print ACE estimation result (both log odds and risk difference)
  print(paste("ATE (Log Odds) for", plot_title, ":", round(ATE, 4)))
  print(paste("ATE (Risk Difference) for", plot_title, ":", round(ATE_rd, 4)))
  
  # Step 6: Check balance using LOVE plot (from cobalt package)
  covariates <- data_clean[, confounders]  # Subset data for covariates
  # bal.tab(covariates, treat = data_clean$CURSMOKE, weights = data_clean$weights, method = "weighting", un = TRUE)  # Balance check
  love.plot(covariates, treat = data_clean$CURSMOKE, weights = data_clean$weights, method = "weighting", binary = "std", threshold = .1)  # LOVE plot
}

# Run the analysis for any confounder set:
estimate_ACE(confounders, "Confounders: EDUC, AGE, SEX, DIABETES")
```

# Model Selection

Because of the DAG we chose SEX, EDUC, AGE as covariates. Here are checking using step-wise model selection and cross-validation.

```{r}
# Step-wise Model Selection

# Select specific columns and create a new dataframe
data_extract <- data[, c("HYPERTEN", "CURSMOKE", "AGE", "EDUC", 
                         "SEX", "BMI", "DIABETES", "HEARTRTE", "RANDID")]


x = as.matrix( data_extract[ , !names(data_extract)%in% c("HYPERTEN","RANDID") ] )
y = data_extract$HYPERTEN

n <- length(y)

set.seed(123)
train_idx <- sample(1:n, size = n*.8)

train_dat <- data_extract[ train_idx, !names(data_extract)%in% c("RANDID")]
test_dat  <- data_extract[-train_idx, !names(data_extract)%in% c("RANDID")]

x_train <- model.matrix(HYPERTEN ~ . -1, train_dat)     
x_test  <- model.matrix(HYPERTEN ~ . -1, test_dat )
y_train <- train_dat$HYPERTEN
y_test  <-  test_dat$HYPERTEN   


# Stepwise selection (AIC/BIC)

# the research question is about 'CURSMOKE', thus we have to use 
# the 'scope' argument to ensure "CURSMOKE" is always included in the model:
# ^2 tells R to also include all possible first-order interactions (in addition to all main effects)
full_model    <- glm(HYPERTEN ~ .^2, data = train_dat, family = binomial)
reduced_model <- glm(HYPERTEN ~ CURSMOKE, data = train_dat, family = binomial)  # Minimum model with CURSMOKE
stepwise_model<- step(full_model, scope = list(lower = reduced_model, upper = full_model), direction = "both")
# scope = list() forces "CURSMOKE" to always stay in the model."CURSMOKE"
# cannot be removed because it’s in the lower bound of model selection.

#summary(stepwise_model)
```

1. best model according to DAG adjustmentSets()

model1 = glm(HYPERTEN ~ CURSMOKE + AGE + DIABETES + EDUC + SEX , family = binomial, data = data)


2. best model according to stepwise model selection ( step() ) and cross-validation ( cv.glmnet() ) without interactions :

model2 = glm(HYPERTEN ~ CURSMOKE + AGE + BMI + HEARTRTE , family = binomial, data = data)


3. Finally, we decide to chose:

model_best = glm(HYPERTEN ~ CURSMOKE + AGE + DIABETES + EDUC + SEX + SEX:AGE , family = binomial, data = data)

# Model Interpretation

```{r}
model_best = glm(HYPERTEN ~ CURSMOKE + AGE + DIABETES + EDUC + SEX + SEX:AGE , family = binomial, data = data)
summary(model_best)
```

1. ATE

```{r}
# codes like those from practical answers of week4 lecture:

data.1 = data
data.1$CURSMOKE = 1

EYhat1 = predict(model_best, newdata = data.1, type = "response")
EY1<-mean(EYhat1,na.rm=T)
# EY1  # 0.3413692

data.0 = data
data.0$CURSMOKE = 0

EYhat0 = predict(model_best, newdata = data.0, type = "response")
EY0<-mean(EYhat0,na.rm=T)
# EY0  # 0.3872214

rd = EY1-EY0 # ATE  -0.04751522 
rr = EY1/EY0
or = EY1*(1-EY0)/(EY0*(1-EY1))
c("E(Y|0)"= EY0, "E(Y|1)"=EY1, "RD"=rd, "RR"=rr, "OR"=or)
```

```{r}
fit.std = stdGlm(fit = model_best, data = data, X = "CURSMOKE")
summary(fit.std)
#ATE 95%CI 
ate.std = summary(fit.std, contrast = 'difference', reference = 0)
ate.std
```

2. ATT

```{r}
# dataset with only the treated individuals
fit.std.treat = stdGlm(fit = model_best, data = data, X = "CURSMOKE", subsetnew = (CURSMOKE == 1) )
summary(fit.std.treat)
att.std = summary(fit.std.treat, contrast = 'difference', reference = 0)
att.std
```

